Sustained software development, not number of citations or journal choice, is indicative of accurate bioinformatic software

### 2021-09-01 manuscript/manuscript-speed-accuracy.tex
Deleting unused figures. Updating README. Renamed supplementey tables.

In the following study, we explore factors that may be indicative of
software accuracy. This, in our opinion, should be one of the prime
reasons for selecting a software tool. We have mined the large and
freely accessible PubMed database \cite{Sayers2010-vm} for benchmarks
of computational biology software, and manually extracted accuracy and
speed rankings for {\color{black}\numTools~} unique software tools. For
each tool, we have collected measures that may be predictive
of accuracy, and may be subjectively employed by the research
community as a proxy for software quality. These include relative
speed, relative age, the productivity and impact of the corresponding
authors, journal impact, number of citations {\color{red}and GitHub activity}.

\section*{Results}
We have collected relative accuracy and speed ranks for
{\color{black}\numTools~} distinct software tools. This software has
been developed for solving a broad cross-section of computational biology
tasks.
Each software tool was benchmarked in at least one of
{\color{black}\numBenchmarkPubs~} publications that satisfy the Boulesteix
criteria \cite{Boulesteix2013-vb}. In brief, the Boulesteix criteria
are: 1. the main focus of the article is a benchmark. 2. the authors
are reasonably neutral. 3. the test data and evaluation criteria are
sensible.

@@

Data on the number of updates to software tools from GitHub such as
the number of versions, commits and contributors was significantly
correlated with software accuracy (respective Spearman’s rhos =
{\color{black}0.15, 0.22, 0.21,} and respective Benjamini \& Hochberg
corrected P-values = {\color{black}0.029, 0.042, 0.063}),
Figure~\ref{fig:allfactors}B.  The significance of these features was
further confirmed with a permutation test
(Figure~\ref{fig:allfactors}B).  These features were not correlated
with speed however (see Figure~\ref{fig:allfactors}A {\color{red}\& Additional file 1: Figures S5\&S6}%\ref{fig:S2}
).  We also found that
reputation metrics such as citations, author and journal H-indices,
and the age of tools were generally \textbf{not} correlated with
either tool accuracy or speed (Figure~\ref{fig:allfactors}A\&B).

@@

In order to gain a deeper understanding of the distribution of
available bioinformatic software tools on a speed versus accuracy
landscape, we ran a permutation test. The ranks extracted
from each benchmark were randomly permuted, generating 1,000
randomized speed and accuracy ranks. In the cells of a $3\times3$
grid spanning the normalised speed and accuracy ranks we computed a
Z-score for the observed number of tools in a cell, compared to the
expected distributions generated by 1,000 randomized ranks. The results
of this are shown in Figure~\ref{fig:speedaccuracy}. We identified {\color{black}4}
of 9 bins where there was a significant excess or dearth of tools. For
example, there was an excess of ``slow and inaccurate'' software ({\color{black}Z=3.39,
P-value=$3.5\times 10^{-4}$}), with more moderate excess of ``slow and accurate'' and ``fast and accurate'' software (Z=2.49 \& 1.7, P=$6.3\times 10^{-3}$ \& 0.04 respectively). We find that only the ``fast and inaccurate'' extreme class is at approximately the
expected proportions based upon the permutation test (Figure~\ref{fig:speedaccuracy}B).

@@

\begin{figure*}[htb!]
%\includegraphics[scale=0.55]{figure2.pdf}
\centering
\includegraphics[width=0.95\textwidth]{figure2.pdf}
\caption{\textbf{A.} A heatmap indicating the relative paucity or abundance of
  software in the range of possible accuracy and speed rankings. Redder
  colours indicate an abundance of software tools in an accuracy and
  speed category, while bluer colours indicate scarcity of software in
  an accuracy and speed category. The abundance is quantified using a
  Z-score computation for each bin, this is derived from 1,000 random
  permutations of speed and accuracy ranks from each
  benchmark. Mean normalised ranks of accuracy and speed have been
  binned into 9 classes (a $3\times 3$ grid) that range from
  comparatively slow and inaccurate to comparatively
  fast and accurate. Z-scores with a P-value less than 0.05 are indicated
  with a ‘$\vartimes$’. \textbf{B.} The z-score distributions from the permutation tests (indicated with the wheat coloured violin plots) compared to 
  the z-score for the observed values for each of the corner and middle square of the heatmap.}
\label{fig:speedaccuracy}
\end{figure*}

@@

We have gathered data on the relative speeds and accuracies of
{\color{black}\numTools~} bioinformatic tools from
{\color{black}\numBenchmarkPubs} benchmarks published between
2005 and 2020. Our results provide significant support for the suggestion that there are major benefits to
the long-term support of software development \cite{siepel2019challenges}.
The
finding of a strong relationship between the number of commits and code contributors to
GitHub (i.e. software updates) and accuracy, highlights the benefits of
long-term or at least intensive development.

@@

\textbf{Criteria for inclusion:} We are interested in using
computational biology benchmarks that satisfy Boulesteix’s
(ALB) three criteria for a ``neutral comparison study''
\cite{Boulesteix2013-vb}. Firstly, the main focus of the article is
the comparison and \textbf{not} the introduction of a new
tool {{\color{red}as these can be biased} \cite{Buchka:2021a}. Secondly, the authors should be reasonably neutral, which means
that the authors should not generally have been involved in the
development of the tools included in the benchmark. Thirdly, the test
data and evaluation criteria should be sensible. This means that the
test data should be independent of data that tools have been trained
upon, and that the evaluation measures appropriately quantify correct
and incorrect predictions.

@@

For stage 2, in order to compute a log-odds score for bioinformatic
words, we computed the frequency of words that were not removed by our
high frequency filter in two different groups of articles:
bioinformatics-background and bioinformatics-benchmark articles. The
text from bioinformatics-background articles were drawn from the
bioinformatics literature, but these were not necessarily associated
with benchmark studies. For background text we used Pubmed
(\cite{Sayers2010-vm,McEntyre2001-fl} to select 8,908 articles that
contained the word “bioinformatics” in the title or abstract and were
published between 2013 and 2015. We computed frequencies for each word
by combining text from titles and abstracts for the background and
training articles. A log-odds score was computed for each word using
the following formula:
%\todo{Word is denoted $w$ on the left and $word$ on the right.}

@@

A score is computed for each candidate article by summing the log-odds
scores for the words in title and abstract,
i.e. $s(a)=\sum_i^Nlo(w_i)$. The high scoring candidate articles are
then manually evaluated against the ALB-criteria. Accuracy and speed
ranks were extracted from the articles that met the criteria, and
these were added to the set of training articles. The evaluated
candidate articles that did not meet the ALB-criteria were incorporated
into the set of background articles. This process was iterated and resulted in the identification of
{\color{black}\numBenchmarkPubs} benchmark articles,
containing {\color{black}\numBenchmarks} different benchmarks. Together these
ranked {\color{black}\numTools~} distinct software packages.

@@

\textbf{Data extraction and processing:} for each article that met the
ALB-criteria and contained data on both the accuracy and speed from
their tests, we extracted ranks for each tool. {\color{red}Until published datasets are made available in consistent, machine-readable formats this step is necessarily a manual process -- ranks were extracted from a mixture of manuscript figures, tables and supplementary materials, each data source is documented in {\color{red} Additional file 2:} Table S1. In addition, a variety of accuracy metrics are reported e.g. ``accuracy'', ``AUROC'', ``F-measure'', ``Gain'', ``MCC'', ``N50'', ``PPV'', ``precision'', ``RMSD'', ``sensitivity'', ``TPR'', ``tree error'', etc. Our analysis makes the necessarily pragmatic assumption that highly ranked tools on one accuracy metric will also be highly ranked on other accuracy metrics.}
Many articles contained
multiple benchmarks, in these cases we recorded ranks from each of these, the
provenance of which is stored with the accuracy metric and raw speed
and accuracy rank data for each tool {\color{red}(Additional file 2: Table S1)}. In line with rank-based
statistics, the cases where tools were tied were resolved by using a
midpoint rank (e.g., if tools ranked 3 and 4 are tied, the rank 3.5 was used)
\cite{Mann1947-re}. Each rank extraction was independently verified by
at least one other co-author to ensure both the provenance of the data
could be established and that the ranks were correct. The ranks for
each benchmark were then normalised to lie between 0 and 1 using the
formula $1-\frac{r-1}{n-1}$ where ‘$r$’ is a tool’s rank and ‘$n$’ is the
number of tools in the benchmark. For tools that were benchmarked
multiple times with multiple metrics (e.g., BWA was evaluated in 6
different articles
\cite{Bao2011-lv,Caboche2014-lj,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Holtgrewe2011-fd})
a mean normalised rank was used to summarise the accuracy and speed performance. 
Or, more formally:

@@

\section*{Acknowledgements}
The authors acknowledge the valued contribution of invaluable
discussions with Anne-Laure Boulesteix, Shinichi Nakagawa, Suetonia
Palmer and Jason Tylianakis. Murray Cox, Raquel
Norel, Alexandros Stamatakis, Jens Stoye, Tandy Warnow, and 
Luis Pedro Coelho provided
valuable feedback on drafts of the manuscript.

>>>>

In the following study, we explore factors that may be indicative of
software accuracy. This, in our opinion, should be one of the prime
reasons for selecting a software tool. We have mined the large and
freely accessible PubMed database \cite{Sayers2010-vm} for benchmarks
of computational biology software, and manually extracted accuracy and
speed rankings for \numTools~ unique software tools. For
each tool, we have collected measures that may be predictive
of accuracy, and may be subjectively employed by the research
community as a proxy for software quality. These include relative
speed, relative age, the productivity and impact of the corresponding
authors, journal impact, number of citations {\color{red}and GitHub activity}.

\section*{Results}
We have collected relative accuracy and speed ranks for
\numTools~ distinct software tools. This software has
been developed for solving a broad cross-section of computational biology
tasks.
Each software tool was benchmarked in at least one of
\numBenchmarkPubs~ publications that satisfy the Boulesteix
criteria \cite{Boulesteix2013-vb}. In brief, the Boulesteix criteria
are: 1. the main focus of the article is a benchmark. 2. the authors
are reasonably neutral. 3. the test data and evaluation criteria are
sensible.

@@

Data on the number of updates to software tools from GitHub such as
the version number, {\color{red}and numbers of contributors, commits, forks and issues } was significantly
correlated with software accuracy (respective Spearman’s rhos =
{\color{red}0.15, 0.21, 0.22, 0.23, 0.23 and respective Benjamini \& Hochberg
  corrected P-values = $5.4\times 10^{-4}$, $1.1\times 10^{-3}$, $7.9\times 10^{-4}$, $3.4\times 10^{-4}$, $3.2\times 10^{-4}$, Additional file 1: Figures S6}). %See arrays "spearmansA" and "spearmansAP" in prettyPlot.R
The significance of these features was
further confirmed with a permutation test
(Figure~\ref{fig:allfactors}B).  These features were not correlated
with speed however (see Figure~\ref{fig:allfactors}A {\color{red}\& Additional file 1: Figures S5\&S6}).
We also found that
reputation metrics such as citations, author and journal H-indices,
and the age of tools were generally \textbf{not} correlated with
either tool accuracy or speed (Figure~\ref{fig:allfactors}A\&B).

@@

In order to gain a deeper understanding of the distribution of
available bioinformatic software tools on a speed versus accuracy
landscape, we ran a permutation test. The ranks extracted
from each benchmark were randomly permuted, generating 1,000
randomized speed and accuracy ranks. In the cells of a $3\times3$
grid spanning the normalised speed and accuracy ranks we computed a
Z-score for the observed number of tools in a cell, compared to the
expected distributions generated by 1,000 randomized ranks. The results
of this are shown in Figure~\ref{fig:speedaccuracy}. We identified 4
of 9 bins where there was a significant excess or dearth of tools. For
example, there was an excess of ``slow and inaccurate'' software (Z=3.39,
P-value=$3.5\times 10^{-4}$), with more moderate excess of ``slow and accurate'' and ``fast and accurate'' software (Z=2.49 \& 1.7, P=$6.3\times 10^{-3}$ \& 0.04 respectively). We find that only the ``fast and inaccurate'' extreme class is at approximately the
expected proportions based upon the permutation test (Figure~\ref{fig:speedaccuracy}B).

@@

\begin{figure*}[htb!]
\centering
\includegraphics[width=0.95\textwidth]{figure2.pdf}
\caption{\textbf{A.} A heatmap indicating the relative paucity or abundance of
  software in the range of possible accuracy and speed rankings. Redder
  colours indicate an abundance of software tools in an accuracy and
  speed category, while bluer colours indicate scarcity of software in
  an accuracy and speed category. The abundance is quantified using a
  Z-score computation for each bin, this is derived from 1,000 random
  permutations of speed and accuracy ranks from each
  benchmark. Mean normalised ranks of accuracy and speed have been
  binned into 9 classes (a $3\times 3$ grid) that range from
  comparatively slow and inaccurate to comparatively
  fast and accurate. Z-scores with a P-value less than 0.05 are indicated
  with a ‘$\vartimes$’. \textbf{B.} The z-score distributions from the permutation tests (indicated with the wheat coloured violin plots) compared to 
  the z-score for the observed values for each of the corner and middle square of the heatmap.}
\label{fig:speedaccuracy}
\end{figure*}

@@

We have gathered data on the relative speeds and accuracies of
\numTools~ bioinformatic tools from
\numBenchmarkPubs~ benchmarks published between
2005 and 2020. Our results provide significant support for the suggestion that there are major benefits to
the long-term support of software development \cite{siepel2019challenges}.
The
finding of a strong relationship between the number of commits and code contributors to
GitHub (i.e. software updates) and accuracy, highlights the benefits of
long-term or at least intensive development.

@@

\textbf{Criteria for inclusion:} We are interested in using
computational biology benchmarks that satisfy Boulesteix’s (ALB) three
criteria for a ``neutral comparison study''
\cite{Boulesteix2013-vb}. Firstly, the main focus of the article is
the comparison and \textbf{not} the introduction of a new tool
{\color{red}as these can be biased} \cite{Buchka:2021a}. Secondly, the
authors should be reasonably neutral, which means that the authors
should not generally have been involved in the development of the
tools included in the benchmark. Thirdly, the test data and evaluation
criteria should be sensible. This means that the test data should be
independent of data that tools have been trained upon, and that the
evaluation measures appropriately quantify correct and incorrect
predictions. {\color{red} In addition, we excluded benchmarks with too
  few tools $\le 3$, or those where the results were inaccessible (no
  supplementary materials or poor figures).}

@@

For stage 2, in order to compute a log-odds score for bioinformatic
words, we computed the frequency of words that were not removed by our
high frequency filter in two different groups of articles:
bioinformatics-background and bioinformatics-benchmark articles. The
text from bioinformatics-background articles were drawn from the
bioinformatics literature, but these were not necessarily associated
with benchmark studies. For background text we used Pubmed
\cite{Sayers2010-vm,McEntyre2001-fl} to select 8,908 articles that
contained the word “bioinformatics” in the title or abstract and were
published between 2013 and 2015. We computed frequencies for each word
by combining text from titles and abstracts for the background and
training articles. A log-odds score was computed for each word using
the following formula:
%\todo{Word is denoted $w$ on the left and $word$ on the right.}

@@

A score is computed for each candidate article by summing the log-odds
scores for the words in title and abstract,
i.e. $s(a)=\sum_i^Nlo(w_i)$. The high scoring candidate articles are
then manually evaluated against the ALB-criteria. Accuracy and speed
ranks were extracted from the articles that met the criteria, and
these were added to the set of training articles. The evaluated
candidate articles that did not meet the ALB-criteria were incorporated
into the set of background articles. This process was iterated and resulted in the identification of
\numBenchmarkPubs benchmark articles,
containing \numBenchmarks~ different benchmarks. Together these
ranked \numTools~ distinct software packages.

@@

\textbf{Data extraction and processing:} for each article that met the
ALB-criteria and contained data on both the accuracy and speed from
their tests, we extracted ranks for each tool. {\color{red}Until
  published datasets are made available in consistent,
  machine-readable formats this step is necessarily a manual process
  -- ranks were extracted from a mixture of manuscript figures, tables
  and supplementary materials, each data source is documented in
 Additional file 2: Table S1. In addition, a variety of
  accuracy metrics are reported e.g. ``accuracy'', ``AUROC'',
  ``F-measure'', ``Gain'', ``MCC'', ``N50'', ``PPV'', ``precision'',
  ``RMSD'', ``sensitivity'', ``TPR'', ``tree error'', etc. Our
  analysis makes the necessarily pragmatic assumption that highly
  ranked tools on one accuracy metric will also be highly ranked on
  other accuracy metrics.}  Many articles contained multiple
benchmarks, in these cases we recorded ranks from each of these, the
provenance of which is stored with the accuracy metric and raw speed
and accuracy rank data for each tool {\color{red}(Additional file 2:
  Table S1)}. In line with rank-based statistics, the cases where
tools were tied were resolved by using a midpoint rank (e.g., if tools
ranked 3 and 4 are tied, the rank 3.5 was used)
\cite{Mann1947-re}. Each rank extraction was independently verified by
at least one other co-author to ensure both the provenance of the data
could be established and that the ranks were correct. The ranks for
each benchmark were then normalised to lie between 0 and 1 using the
formula $1-\frac{r-1}{n-1}$ where ‘$r$’ is a tool’s rank and ‘$n$’ is
the number of tools in the benchmark. For tools that were benchmarked
multiple times with multiple metrics (e.g., BWA was evaluated in 6
different articles
\cite{Bao2011-lv,Caboche2014-lj,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Holtgrewe2011-fd})
a mean normalised rank was used to summarise the accuracy and speed
performance.  Or, more formally:

@@

\section*{Acknowledgements}
The authors acknowledge the valued contribution of invaluable
discussions with Anne-Laure Boulesteix, Shinichi Nakagawa, Suetonia
Palmer and Jason Tylianakis. Murray Cox, Raquel
Norel, Alexandros Stamatakis, Jens Stoye, Tandy Warnow, and 
{\color{red}Luis Pedro Coelho and three anonymous reviewers} provided
valuable feedback on drafts of the manuscript.
